Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-22 paper/main.tex
trying to explain sp selection

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors; but here we were particularly concerned because
the {\it de novo} gene caller used by the references was Augustus for many species 
(e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; {\it Asparagus officinalis}, 
\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and a different HMM-based tool for many more (e.g. {\it Citrus clementina}, \cite{wu2018genomics}; 
{\it Capsella grandiflora}, \cite{slotte2013capsella};,...), which would potentially bias 
reference based comparisons of Helixer's and Augustus' performance. % examples chosen alphabetically and skipping ons where I wasn't sure from the methods

We therefore downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{} % supp phylogenies maybe?
within our generalizable range (i.e. excluding algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, particularly to
exonic regions; that had low 3' bias, normal FastQC and Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.

>>>>

\subsection{Evaluation against independent RNAseq data}
\label{sec:rnaseq}
The reference annotations were created with existing tools and largely with a pipeline incorporating
{\it de novo} gene predictions with RNAseq and homology data. The references, like any
data, are expected to contain errors so we chose to use RNAseq data for independent evaluation.
In plants there is the additional concern that the references may share biases with
Augustus or HMMs as Augustus was used as the {\it de novo} gene caller for the reference of
many plant species (e.g. {\it Arabidopsis lyrata}, \cite{rawat2015improving}; 
{\it Asparagus officinalis}, 
\cite{harkess2017asparagus}; and {\it Chenopodium quinoa}, \cite{jarvis2017genome})  
and other HMM-based tools for many more (e.g. {\it Citrus clementina}, \cite{wu2018genomics}; and
{\it Capsella grandiflora}, \cite{slotte2013capsella}). The Ensemble animal dataset by and large
used the Ensembl annotation pipeline, which includes no HMM-based tool.

We downloaded and processed public RNAseq data to obtain an independent option 
for evaluating model performance. We selected three each of plant and animal genomes for
detailed evaluation with RNAseq. These were selected to have relatively good 
({\it Manihot esculenta} and {\it Papio anubis}), 
typical ({\it Medicago truncatula} and {\it Equus caballus}), 
and poor ({\it Theobroma cacao} and {\it Petromyzon marinus})
performance compared to Augustus \ref{} % supp phylogenies maybe?
within our generalizable range (i.e. excluding algae and
invertebrates). Selections were further constrained by the availability of stranded RNAseq
data.

% todo, add versions to everything below
For each of these six species the following search was performed on Sequence Read Archive
'(("<species name>"[Organism] OR <species name>[All Fields]) AND stranded[All Fields]) 
AND ("biomol rna"[Properties] AND "library layout paired"[Properties])'. In so far as 
more than 50 samples were identified, every Nth sample was selected so that about 50
samples were chosen for further processing, as listed here \ref{}. % todo make a supp table
Each sample was quality controlled with FastQC \citep{}, before and after trimming
with Trimmomatic \citep{bolger2014trimmomatic}, and then mapped to the genome with Hisat2 \citep{kim2019graph}. 
Samtools \citep{li2009sequence} was used for the sorting and indexing that was necessary both for
RNAseq mapped metric collection with PicardTools \citep{Picard2019toolkit} and custom downstream 
processing below. MultiQC \citep{ewels2016multiqc} was used to summarize the results. The above was 
automated and the results visualized with the code here (... make available and link to RNAseq stuff).
Samples were filtered to those that had {\it relatively} high mapping rates, high mapping to
exonic regions relative to other regions; that had low 3' bias, normal FastQC and 
Trimmomatic stats, and that
were confirmed to be stranded with 2nd strand as sense strand. If more remained, seven
of the high quality samples were selected randomly.
